---
title: "Introduction to R"
author: "Koo Ping Shung"
date: "5 October 2016"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

##Welcome

Welcome to Introduction to R. Below are the learning objectives of the course

- Understand how R works
- Perform simple computation using R
- Know what are the variables in R and how to convert them
- Perform Conditional Processing and Computation Loops

##What is R?
R is an open-source language with multiple libraries(pacakages) available, that allows the user to undertake various tasks such as computation, data wrangling, visualization and so on. These packages are mostly user-created allowing for more sophisticated computation. 

##Doing Calculations
R like any other programming language that you come across can provide simple calculation. For instance, 

```{r, eval = F,echo=TRUE}
12 + 1 #addition
9 - 4 #substraction
23 + 2 - 4 #addition & substraction
112 * 2 #multiplication
84 / 14 #division
2^4 #power
```

##Data Types in R

There are several data types in R

- Data.Frame
- Matrices
- Vectors
- Characters
- Numeric & Integer
- Logical
- Factor
- Date & Time





