---
title: "Introduction to R"
author: "Koo Ping Shung"
date: "5 October 2016"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

##Welcome

Welcome to Introduction to R. Below are the learning objectives of the course

- Understand how R works
- Perform simple computation using R
- Know what are the variables in R and how to convert them
- Perform Conditional Processing and Computation Loops

##What is R?
R is an open-source language with multiple libraries(pacakages) available, that allows the user to undertake various tasks such as computation, data wrangling, visualization and so on. These packages are mostly user-created allowing for more sophisticated computation. 

##Doing Calculations
R like any other programming language that you come across can provide simple calculation. For instance, 

```{r, eval = F,echo=TRUE}
12 + 1 #addition
9 - 4 #substraction
23 + 2 - 4 #addition & substraction
112 * 2 #multiplication
84 / 14 #division
2^4 #power
```

##Data Types in R

There are several data types in R

- Numeric & Integer
- Characters
- Logical
- Factor
- Date & Time
- Data.Frame
- Matrices
- Vectors

Let's look at them in more details.

##Numeric & Interger
- In R, numeric and integer are different.
- Main difference as follows:

Properties    | Numeric      | Integer
:-------------|:------------:|:-------------: 
Whole Number  | Unnecessary  | Necessary     
Decimal Place | Yes          | No
Example       | 2.04         | 4

- Numbers in R are treated as numeric unless specified.
- To explicit state an integer you need to specify an *L suffix*

```{r,eval=T,echo=TRUE}
class(5L)
```

##Characters
* Also known as *'strings'*
* Can contain any character, symbol or numbers
* Comprise of pure or mixture of any of the stated values
* For example, 
    * aabbcdcbabcdbabcdb (pure)
    * 127781827 (pure)
    * ##&&*@@ (pure)
    * John gave Alice his email address as "John@hotmail.com." (mixture)

##Logical
- usually created when we want to make comparison
- Consist of only two value "TRUE" or "FALSE"

For example,

```{r, eval=TRUE,echo=TRUE}
x=3;y=1
z=x>y
z

class(z)
```

##Data.Frames

- Used for storing tabular data
- Can be seen as a list of vectors of equal length
- Each row in the data.frame represent an observation and each columns represent an attribute (also known as variable)





