---
title: "Introduction to R"
author: "Koo Ping Shung"
date: "5 October 2016"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

##Welcome

Welcome to Introduction to R. Below are the learning objectives of the course

- Understand how R works
- Perform simple computation using R
- Know what are the variables in R and how to convert them
- Perform Conditional Processing and Computation Loops

##What is R?
R is an open-source language with multiple libraries(pacakages) available, that allows the user to undertake various tasks such as computation, data wrangling, visualization and so on. These packages are mostly user-created allowing for more sophisticated computation. 

##Doing Calculations
R like any other programming language that you come across can provide simple calculation. For instance, 

```{r, eval = F,echo=TRUE}
12 + 1 #addition
9 - 4 #substraction
23 + 2 - 4 #addition & substraction
112 * 2 #multiplication
84 / 14 #division
2^4 #power
```

##Data Types in R

There are several data types in R

- Numeric & Integer
- Characters
- Logical
- Factor
- Date & Time
- Data.Frame
- Matrices
- Vectors
- Lists

Let's look at them in more details.

##Numeric & Interger
- In R, numeric and integer are different.
- Main difference as follows:

Properties    | Numeric      | Integer
:-------------|:------------:|:-------------: 
Whole Number  | Unnecessary  | Necessary     
Decimal Place | Yes          | No
Example       | 2.04         | 4

- Numbers in R are treated as numeric unless specified.
- To explicit state an integer you need to specify an *L suffix*

```{r,eval=T,echo=TRUE}
class(5L)
```

##Special Values

- There are 2 special values under numeric/integer.
- Special Value 1: inf
    - This value represents *infinity*.
- Special value 2: NaN
    - This represents *undefined value*, also associated with *missing value*.
```{r, eval=TRUE,echo=TRUE}    
class(Inf)

class(NaN)
```

##Characters
* Also known as *'strings'*
* Can contain any character, symbol or numbers
* Comprise of pure or mixture of any of the stated values
* For example, 
    * aabbcdcbabcdbabcdb (pure)
    * 127781827 (pure)
    * ##&&*@@ (pure)
    * John gave Alice his email address as "John@hotmail.com." (mixture)

##Logical
- usually created when we want to make comparison
- Consist of only two value "TRUE" or "FALSE"

For example,

```{r, eval=TRUE,echo=TRUE}
x=3;y=1
z <- x>y
z

class(z)
```
 
##Factor
- Factors are categorical variables
- It need not be ordered like "Small", "Medium", "Large"
- For example,
    - Gender
    - Branch Codes
    - T-Shirt Sizes

##Date

- Like most software, dates in R is stored as the number of days between the date value and January 1st, 1970.
- For instance, to store "1970-01-01", R would store it as "0".
- To store "1971-01-01", R would store it as "365".
- Thus dates **before** reference date will reflect **negative value**, while dates **after** reference date will reflect **positive** value.

##Time

- Time is also referenced to January 1st, 1970.
- Time can be stored in either class:
    - POSIXct(): data stored in the form of seconds (default)
    - POSIXlt(): data sored in list of elements

```{r, eval=TRUE,echo=TRUE}    
b <- as.POSIXlt(Sys.time())
b
b$min
```

##Time Element

Elements   | Range   |Description
:---------:|:-------:|:----------
sec        |0-61     |Seconds
min        |0-59     |Minutes
hour       |0-23     |Hours
mday       |1-31     |Day of Month
mon        |0-11     |Months after first month of year
year       |>1900    |Year after 1900
wday       |0-6      |Day of Week, "0" represent Sunday
yday       |0-365    |Day of the year
isdst      |0-1      |Daylight saving time flag

##Data.Frames

- Used for storing tabular data
- Can be seen as a list of vectors of equal length
- Each row in the data.frame represent an observation and each columns represent an attribute (also known as variable)

```{r,eval=T,echo=T}
class(iris) #iris is an existing dataset in R
```
- We shall discuss this in more details later

##Vector
- Sequence of data of the same type
- Members in a vector are called components (or just members)
- Implicit meaning usually attached to the position of each component
- **c()** is used to create the vector as seen in the example below

```{r,eval=T,echo=T}
x <- c(1.4, 5.7, 3.2, 4.3, 9.1)
x
```

##Matrix
- A collection of data arranged in 2 dimension layout
- An array with 2 subscripts
- Forms the most fundamental of data science
```{r,eval=T,echo=T}
y <- matrix(c(1.4, 5.7, 3.2, 4.3, 9.1, 7.8),nrow=2,ncol=3)
y
```

##Lists
- Special type of vector that can contain elements of different type.
```{r,eval=T,echo=T}
x <- list(4,"TRUE", TRUE, c(1,4,7.3))
class(x)
class(x[[2]])
class(x[[3]])
class(x[[4]])
```

##Attributes
- Each datatypes can have certain attributes.
- To understand more of each attributes, we can use the **attributes()** function.
```{r,eval=T,echo=T}
x <- list(4,"TRUE", TRUE, c(1,4,7.3))
attributes(x)
y <- matrix(c(1.4, 5.7, 3.2, 4.3, 9.1, 7.8),nrow=2,ncol=3)
attributes(y)
```

##Assignments
- By now you would have seen how you can assign an expression to an object in R
- "<-" is the assignment operator
- For example, I am assigning the expression "4+1"" to x
- You'll notice that the expression is evaluated before assigning the final result to x

```{r,eval=T,echo=T}
x <- 4+1
x
```
- If you do not want the R to evaluate the expression then you can use **quote()**

```{r,eval=T,echo=T}
y <- quote(9+13)
y
```

##Explicit Coercion
- Sometimes for statistical modelling to work, you may need to change the data type of a column
- To do that, we use *explicit coercion*
- Generally, you just have to select the column/object and use as.*() to change
- For example,
```{r,eval=T,echo=T}
x <- c(0,1,1,0,0)
class(x)
x <- as.factor(x)
class(x)
```

##Comparison
- Often times, you wanted to check if a certain element is of certain type. You can do the following:
```{r,eval=T,echo=T}
x <- TRUE
u <- is.character(x)
u
v <- is.factor(x)
v
w <- is.logical(x)
w
```
- To compare if an object is of a certain type/class, you just need is.*() to make the comparison.


##Subsetting
- At anytime, you may only need a subset of a data type
- Taking a subset depends on the data type.
- "[]" returns an object the same class as the original
- "[[]]" used to extract elements of a list or data.frames. Onced use, the return value need not necessary be the same type as the list or data.frames.
- "$" is used to extract elements of a data.frames or list. Often times, we use it to refer to columns in data.frames. 

Some examples as follows:

##Subsetting Examples
```{r,eval=T,echo=T}
x <- c(9,2,5,7,8,4,3,1)
x[4]
x[2]
x[x>5]
```

##More Subsetting Examples
```{r,eval=T,echo=T}
x <- c(9,2,5,7,8,4,3,1)
u <- x>4
u
```

##Subsetting Matrix
- Let's set up a matrix for first
```{r,eval=T,echo=T}
x <- matrix(c(9,2,5,7,8,4,3,1,6,10,19,13),nrow=3,ncol=4)
x
```
-If I want the element on 2nd row and 3rd column, I will need to code as
```{r,eval=T,echo=T}
x[2,3]
```

##Subsetting Matrix (continued)
- If I want all the elements on 1st row, I will need to code as
```{r,eval=T,echo=T}
x[1,]
```
- How about if I want all the element on 4th column? Here you go.
```{r,eval=T,echo=T}
x[,4]
```
- Let's make it more challenging. I want 2nd and 3rd column.
```{r,eval=T,echo=T}
x[,2:3]
```

##Subsetting Matrix (Continued)
- You will notice so far, if you subset by rows or columns they return in a vector of length 1
- If that is not what you wanted, try *drop=FALSE* option.
```{r,eval=T,echo=T}
x[,4,drop=FALSE]
```

##Random Number Generator
- Often times in Data Science work, you might need to generate a set of numbers
- For instance, if I want to generate a set of 13 numbers with mean=4 and standard deviation=3
```{r,eval=T,echo=T}
x<-rnorm(13,4,3)
x

summary(x)
```
The above is generating from a Normal probability distribution. For other probability distribution, you may Google for them easily.

##Random Sample
- The *sample()* function can be used to select a random sample from a set of (scalar) objects.
```{r,eval=T,echo=T}
set.seed(123789)
u<-sample(1:30,10)
u

v<-sample(1:20,10,replace = TRUE)
v
```

##Functions
- Similar to Excel functions that you are familiar with
- Can be pre-defined in R packages or user-created
- Define the process that transforms or extracts information from data values
- Having functions allows
    - reduces the code to maintain
    - readability
    - saves programmer's time
- To use functions, we need to know the function's name and their arguments
- General form of function is as follows
    
    *function-name(argument1, argument2, argument3,...argument n)*
    
##Matching Arguments
- To avoid confusion on the position of argument, state the name of the argument in the function 
- For example,

```{r,eval=T,echo=T}
m <- matrix(c(4,1,6,8,3,9),nrow = 2,ncol=3) #is similar to below
n <- matrix(c(4,1,6,8,3,9),ncol = 3,nrow=2)
m
n
```

##User Created Function
- User can create their own functions too
- Check out the following

```{r,eval=T,echo=T}
testf <- function(a,b,c){
  (a*b)-c
}
g <- testf(9,10,14)
g
```

##Unmatched Arguments

- Going back to the testf() function, let's try the following
```{r,eval=F,echo=T}
f <- testf(14,10)
```

You would see an error that says an argument is missing. This could be avoided by assigning default values.

```{r,eval=T,echo=T}
testf <- function(a=1,b=1,c=0){
  (a*b)-c
}
g <- testf(9,10)
g
```

##Control Structures
- There are times we want some of our functions to generate output given certain conditions.
- In the next few slides we will go through the following: 
    - Conditional Processing
    - Loop & Nested Loop
    - Control Structure
        - While
        - Repeat & Break
        - Next 
        - Return

##Conditional Processing
- To process data according to certain conditions
- Example,

```{r,eval=T,echo=T}
x <- 4
if(x > 5) {
	y <- "Greater than five"
} else {
	y <- "Less than or equal to 5"
}
y
```

##Operators

Symbols  |Description
---------|-------------
   "=="  |Equality
   ">"   |Greater than
   "<"   |Less than
   ">="  |Greater than or Equal to
   "<="  |Less than or Equal to
   "&&"  |AND
   "||"  |OR
   "!"   |NOT
   
##Loops

- Create loops to process data iteratively
- Useful for cleaning data

```{r,eval=T,echo=T}
y <-c(1,5,7,3,2,7,8,9,3)
x <- vector("logical", length = 8)
for(i in 1: length(y)) {
  x[i] <- y[i] > 5
}
x
```

##Nested Loops
- Known as loop within loop
```{r,eval=T,echo=T}
trymatrix <- matrix(nrow = 12, ncol = 12)
for(i in 1:dim(trymatrix)[1]) { #outer loop
			for(j in 1:dim(trymatrix)[2]) { #inner loop
				trymatrix[i, j] <- i*j
			}
		}
trymatrix
```

##While Loop
- Tells R to keep processing as long as condition is satisfied
```{r,eval=T,echo=T}
x <- 3
	while(x <= 33) {
		x <- x + 3
		print(x)
}
```

##Repeat Loop & Break Clause
- Repeat loop and "break" clause works together
- Repeat loop commonly used when conditions need to be "repeated" (unlike while loop, which stops when condition not met)
- An instruction within a repeat loop will be executed at least once
- To prevent an infinite loop when using repeat loop, can use if.else with break clause
```{r,eval=T,echo=T}
x <- 2
	repeat {
	  x <- x + 1
	  print(x)
		if(x == 10) {
		break
		}
	}
```

##Next Clause
- Next clause tells R to skip an iteration of the loop
```{r,eval=T,echo=T}
c1 <- 0
c2 <- 0
for(i in 1:100) {
			c1 <- c1 + 1
			if(i <= 15) {
				next
			}
			c2 <- c2 + 1
		}
c1
c2
```

##Return Clause
- Tells R the type of value to return
- Particularly used in user-defined functions
```{r,eval=T,echo=T}
startT <- function(x){
		if('T' == substr(x,1,1)){
		return(TRUE)
		}
		return(FALSE)
    }
t1 <- startT("The cow jump over the moon")
t1
```

##lapply
- Often times you want to apply same function over a list of items
- lapply() does that
    - First argument for lapply is a list
    - Second argument is a name of a function
    - Third argument are other arguments for the function
- For example,
```{r,eval=T,echo=T}
x <- list(a=rnorm(10), b=10:50)
y <- lapply(x, mean)
y
```


##Install Packages
- In R, you may need to install R packages for your work
- Packages may contain relevant functions or data.frames that you need
- For example, you want to use the *dplyr* package. Submit the following code
```{r,eval=F,echo=T}
install.packages("dplyr")
```
- After installation, for any new R session, you need to deploy the packages you need. To deploy, 
```{r,eval=F,echo=T}
library(dplyr)
```

##Coding Guidelines
- This wraps up the introduction to R workshop
- As you step into the world of coding, we would like to provide some coding guidelines

##Why Coding Guidelines?
- To make codes more comprehensible, not only for you but others as well
- Understand the process flow of a code easily
- State the boundaries of the individual codes, especially loops and user-defined functions visually
    - Useful when reviewing and debugging codes
- All that said, guidelines are not mandatory (we strongly recommend though)

##Coding Guidelines: Consolidating
- Consolidate all similar codes in a single program
- Done in a text editor, to test out the codes
- *Important*!
    - Specify deployed packages before the codes
    - Codes that manage R session should be placed at the start of the program

##Coding Guidelines: Commenting
- Comment on codes where possible
- Help viewers understand the code better
- Easier to review codes
- To comment in R, start with a "#" sign then follow by the comment

##Coding Guidelines: Indentation
- Defines the boundary of codes
- For example, 
```{r,eval=F,echo=T}
for(i in 1:dim(trymatrix)[1] {
		for(j in 1:dim(trymatrix)[2] {
			trymatrix[i, j] <- i*j
		}
	}

```
- With the above example, other programmers can define the boundaries of the loops easily


   
